---
title: 'DATA 605: Computational Mathematics Discussion Board 13'
author: "Gabriel Campos"
date: "Last edited `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
  geometry: left=0.5cm,right=0.5cm,top=1cm,bottom=2cm
  pdf_document:
    latex_engine: xelatex
  html_notebook: default
urlcolor: blue
---

# Week 13 Discussion:

Using R, provide the solution for any exercise in either Chapter 4 or Chapter 7 of the calculus textbook. If you are unsure of your solution, post your concerns.

Page 179 Exercise 4.2 Question 3

Water flows onto a flat surface at a rate of $5\frac{cm^3}{s}$ forming a circular puddle $10mm$ deep. How fast is the radius growing when the radius is: (a) 1 cm?  (b) 10 cm?  (c) 100 cm?

First we just need to define our variables:

- We can express $\frac{dV}{dt}=5\frac{cm^3}{s}$ as the rate of the puddle increasing
- The water flow surface/ circular puddle describes a cylinder. Formula for a cylinder is $Volume=Area \ base \ \times \ Height$
- $Area \ Base_{circular} = \pi r^2$
- Height is given as $10mm$ and when converted to centimeters it becomes $1 \ cm$
- $\therefore \ V=\pi r^2 (1)$
- We can use the derivative on the Volume formula $V=\pi h r^2$ making $\frac{dV}{dt}=\frac{d}{dt}(\pi r^2)$ 
- which gives us $\frac{dV}{dt}=2 \pi r\frac{dr}{dt}$
- Simplifying this we get: $\frac{\frac{dV}{dt}}{2 \pi r } = \frac{dr}{dt}$ 
- Substitute the rate to get $\frac{5}{2 \pi r } = \frac{dr}{dt}$

**How fast is the radius growing** is defined by $\frac{dr}{dt}$ therefore this is what we are looking for.

our height is given for sections a - c.

(a) 1 cm? $\frac{5}{2 \pi r_{\ 1 \ cm}}$ $\rightarrow$ $\frac{5}{2 \pi (1)}$

```{r}
5/(2*pi*1)
```

(b) 10 cm? $\frac{5}{2 \pi r_{\ 10 \ cm}}$ $\rightarrow$ $\frac{5}{2 \pi (10)}$

```{r}
5/(2*pi*10)
```

(c) 100 cm? $\frac{5}{2 \pi r_{\ 100 \ cm}}$ $\rightarrow$ $\frac{5}{2 \pi (100)}$

```{r}
5/(2*pi*100)
```
