---
title: 'DATA 605: Computational Mathematics Discussion Board 12'
author: "Gabriel Campos"
date: "Last edited `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
  geometry: left=0.5cm,right=0.5cm,top=1cm,bottom=2cm
  pdf_document:
    latex_engine: xelatex
  html_notebook: default
urlcolor: blue
---

# Library

```{r, message=FALSE}
library(dplyr)
```

# Assignment

Using R, build a multiple regression model for data that interests you.  Include in this model at least one quadratic term, one dichotomous term, and one dichotomous vs. quantitative interaction term.  Interpret all coefficients. Conduct residual analysis.  Was the linear model appropriate? Why or why not?

In this course, do not attach files to any of the discussion posts.  Provide your code and commentary plus figures directly in the text box.  Many students don't download files or open other links.

# Data

[Kaggle Medical Cost Personal Datasets](https://www.kaggle.com/datasets/mirichoi0218/insurance)

```{r}
df_ins <- read.csv("insurance.csv")
```

```{r}
dim(df_ins)
```

```{r}
sum(is.na(df_ins))
```

# Quadratic

$ax^2+bx+c=0$

Ill make a quadratic using the bmi column which allows the model to capture any non-linear relationship between bmi and charges.

```{r}
df_ins$quadratic<-df_ins$bmi^2
```

# Dichotomous

Dichotomous are nominal variables with 2 levels or categories.

Gender is categorized this way, therefore it will be multiplied by the quadratic variable bmi. In doing so I can assess whether the relationship between bmi and charges differs by gender.

```{r}
df_ins$gender_binary <- ifelse(df_ins$sex == "male", 1, 0)
df_ins$dichotomous<-df_ins$bmi*df_ins$gender_binary
```


# Linear Model

```{r}
lm <- lm(charges ~ gender_binary + dichotomous + bmi + children + smoker + region + age,data = df_ins)
summary(lm)
```

```{r}
residual<- resid(lm)
plot(residual, col = 'dark red')

```


```{r}
hist(residual, col = "steelblue")
```

```{r}
qqnorm(residual, col = 'steelblue')
qqline(residual)
```


The result show:

- $R^2 = 0.7509$ indicates $\approx 75%$  of variance is captured which makes this a good model with high statistical relevance.
- bmi coefficient of 336.88 means that for each additional increase in bmi, charges will increase approximately by $336.88
- the low p-value for bmi indicates it is highly statistically significant, making it a strong predictor.
- dichotomous is 4.59, representing the interaction effect between bmi and gender_binary. This indicates that for each unit increase of bmi, the charges between males (gender_binary=1) and females (gender_binary=0) is expected to increase by 4.59.
- the high p-value of $\approx 90.3$ means it is not statistically significant.